---
title: "Exploratory Data Analysis"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    fig-width: 9
    fig-height: 6
---

```{r}

#| label: setup
#| include: false

options(java.parameters = "--add-opens=java.base/java.nio=ALL-UNNAMED")
source("/Users/cepickering/Documents/Large Folders/Git/uci_diabetes/reports/part1/R/connections.R")

library(tableone)
library(kableExtra)
library(naniar)
library(skimr)
library(patchwork)
library(scales)
library(caret)

```

## Overview

This section presents an exploratory analysis of the UCI 130-US Hospitals 
Diabetes dataset following preprocessing and feature engineering in dbt. 
The analytic sample consists of **69,584 encounters** after deduplication 
to one encounter per patient, removal of deceased and hospice discharge 
dispositions, neonatal admissions, and encounters with missing primary diagnosis.
The primary outcome is **30-day hospital readmission** (binary).

## Outcome Distribution and Class Imbalance

```{r}

#| label: outcome-distribution

outcome_counts <- df |>
  count(readmitted_30day) |>
  mutate(
    label = ifelse(readmitted_30day == 1, "Readmitted <30 days", "Not readmitted"),
    pct = n / sum(n),
    pct_label = paste0(label, "\n", comma(n), " (", percent(pct, accuracy = 0.1), ")")
  )

ggplot(outcome_counts, aes(x = label, y = n, fill = factor(readmitted_30day))) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = paste0(comma(n), "\n(", percent(pct, accuracy = 0.1), ")")),
            vjust = -0.3, size = 4) +
  scale_fill_manual(values = c("#4E9BB9", "#C0392B")) +
  scale_y_continuous(labels = comma, limits = c(0, max(outcome_counts$n) * 1.15)) +
  labs(
    title = "30-Day Readmission Outcome Distribution",
    subtitle = paste0("Positive class rate: ",
                      percent(mean(df$readmitted_30day == 1), accuracy = 0.1)),
    x = NULL,
    y = "Number of Encounters"
  ) +
  theme_classic() +
  theme(
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 12),
        axis.text.x = element_text(size = 13),
        axis.text.y = element_text(size = 10)
  )

```

## Missingness Profile

```{r}

#| label: missingness-table

miss_summary <- df |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  mutate(pct_missing = percent(n_missing / nrow(df), accuracy = 0.01)) |>
  arrange(desc(n_missing))

miss_summary |>
  kable(
    col.names = c("Variable", "N", "%"),
    caption = "Variables with Missing Values"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

```{r}

df |>
  miss_var_summary() |>
  filter(n_miss > 0) |>
  ggplot(aes(x = reorder(variable, pct_miss), y = pct_miss)) +
  geom_col(fill = "#4E9BB9") +
  coord_flip() +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Missingness by Variable",
    subtitle = "Variables with at least one missing value",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 12),
        axis.text.x = element_text(size = 13),
        axis.text.y = element_text(size = 13)
  )

```

```{r}

missingness_decisions <- tibble::tibble(
  variable = c("medical_specialty", "race", "diag_3", "diag_2", "gender"),
  pct_missing = c("48.12%", "2.69%", "1.64%", "0.41%", "0.00%"),
  mechanism = c(
    "High missingness — likely MNAR (not documented for certain specialties)",
    "Likely MCAR — small, no clear pattern",
    "Structural — not all admissions have tertiary diagnosis",
    "Structural — not all admissions have secondary diagnosis",
    "Negligible — 3 records, likely data entry error"
  ),
  decision = c(
    "Retain 'Unknown' as explicit category — already handled in specialty_group",
    "Mode imputation; may want a more robust approach in Part 2 preprocessing",
    "Already handled — NULL maps to 'None' in diag_3_group",
    "Already handled — NULL maps to 'None' in diag_2_group",
    "Drop 3 records in Part 2 preprocessing — negligible loss"
  )
)

missingness_decisions |>
  kable(
    col.names = c("Variable", "% Missing", "Mechanism", "Handling Decision"),
    caption = "Missingness Assessment and Handling Decisions"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

## Table 1

```{r}

#| label: table-one

# Verify column names match exactly
# names(df)  # uncomment to inspect if needed

# Define variable lists — all lowercase per janitor::clean_names()
cat_vars <- c(
  "race", "gender", "age_band", "age_group",
  "admission_type_group", "specialty_group",
  "diag_1_group", "diag_2_group", "diag_3_group",
  "max_glu_serum", "a1cresult",
  "metformin", "repaglinide", "nateglinide", "chlorpropamide",
  "glimepiride", "glipizide", "glyburide", "tolbutamide",
  "pioglitazone", "rosiglitazone", "acarbose", "miglitol",
  "tolazamide", "insulin", "glyburide_metformin",
  "glipizide_metformin", "glimepiride_pioglitazone",
  "metformin_rosiglitazone", "metformin_pioglitazone",
  "change", "diabetes_med",
  "readmitted_raw", "split_group"
)

cont_vars <- c(
  "age_midpoint", "time_in_hospital",
  "num_lab_procedures", "num_procedures", "num_medications",
  "number_outpatient", "number_emergency", "number_inpatient",
  "number_diagnoses", "total_prior_encounters",
  "num_med_changes", "num_meds_active"
)

# Prepare data — factor the strata variable with labels
df_tab <- df |>
  mutate(
    readmitted_30day = factor(
      readmitted_30day,
      levels = c(0, 1),
      labels = c("Not Readmitted", "Readm <30d")
    ),
    # Cast continuous vars explicitly
    across(all_of(cont_vars), as.numeric),
    # Factor ordering for key categoricals
    age_group = factor(age_group, levels = c("young", "middle", "senior")),
    age_band = factor(age_band, levels = c(
      "[0-10)", "[10-20)", "[20-30)", "[30-40)", "[40-50)",
      "[50-60)", "[60-70)", "[70-80)", "[80-90)", "[90-100)"
    )),
    admission_type_group = factor(admission_type_group,
      levels = c("Emergency", "Elective", "Newborn", "Unknown")),
    readmitted_raw = factor(readmitted_raw,
      levels = c("NO", ">30", "<30")),
    split_group = factor(split_group,
      levels = c("train", "validation", "holdout")),
    # Medication columns — logical ordering
    across(all_of(c(
      "metformin", "repaglinide", "nateglinide", "chlorpropamide",
      "glimepiride", "glipizide", "glyburide", "tolbutamide",
      "pioglitazone", "rosiglitazone", "acarbose", "miglitol",
      "tolazamide", "insulin", "glyburide_metformin",
      "glipizide_metformin", "glimepiride_pioglitazone",
      "metformin_rosiglitazone", "metformin_pioglitazone",
      "change", "diabetes_med"
    )), ~ factor(., levels = c("No", "Steady", "Up", "Down"))),
    # Diag groups — alphabetical is fine, Diabetes first makes clinical sense
    diag_1_group = factor(diag_1_group, levels = c(
      "Diabetes", "Circulatory", "Respiratory", "Digestive",
      "Genitourinary", "Injury", "Musculoskeletal", "Neoplasms", "Other"
    )),
    diag_2_group = factor(diag_2_group, levels = c(
      "None", "Diabetes", "Circulatory", "Respiratory", "Digestive",
      "Genitourinary", "Injury", "Musculoskeletal", "Neoplasms", "Other"
    )),
    diag_3_group = factor(diag_3_group, levels = c(
      "None", "Diabetes", "Circulatory", "Respiratory", "Digestive",
      "Genitourinary", "Injury", "Musculoskeletal", "Neoplasms", "Other"
    )),
    max_glu_serum = factor(max_glu_serum,
      levels = c("None", "Norm", ">200", ">300")),
    a1cresult = factor(a1cresult,
      levels = c("None", "Norm", ">7", ">8")),
    race = factor(race, levels = c(
      "Caucasian", "AfricanAmerican", "Hispanic", "Asian", "Other"
    )),
    gender = factor(gender, levels = c("Male", "Female"))
  )

```

```{r}

#| label: table-one-render
#| 
# Create Table 1 using tableone
tab1 <- CreateTableOne(
  vars = c(cat_vars, cont_vars),
  strata = "readmitted_30day",
  data = df_tab,
  factorVars = cat_vars,
  addOverall = TRUE
)

tab1_df <- print(
  tab1,
  showAllLevels = FALSE,
  nonnormal = cont_vars,
  quote = FALSE,
  noSpaces = TRUE,
  printToggle = FALSE
)

# Identify header rows (not indented with 3 spaces)
header_rows <- which(!grepl("^   ", rownames(tab1_df)))

# Restore indentation in row names before passing to kable
rownames(tab1_df) <- ifelse(
  grepl("^   ", rownames(tab1_df)),
  paste0("\u00a0\u00a0\u00a0\u00a0\u00a0\u00a0", trimws(rownames(tab1_df))),
  rownames(tab1_df)
)

tab1_df |>
  kable(
    format = "html",
    escape = FALSE,
    caption = "Table 1: Patient Characteristics Stratified by 30-Day Readmission"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    font_size = 12,
    full_width = TRUE
  ) |>
  row_spec(header_rows, extra_css = "border-top: 2px solid #929393ff;") |>
  scroll_box(height = "800px")

```

## Near-zero variance flagging

Seven variables have zero variance and therefore carry no information. Another six variables 
have >95% of values in a single category, which may cause modeling issues and should be considered 
for removal or re-categorization. The table below summarizes these near-zero variance variables.
Repaglinide and max_glu_serum may be retained for modeling, but the others should not be. We will reassess later.

```{r}

#| label: near-zero-variance

nzv_results <- df |>
  select(all_of(cat_vars)) |>
  mutate(across(everything(), as.character)) |>
  summarise(across(everything(), ~ {
    tbl <- table(.)
    max(tbl) / sum(tbl)
  })) |>
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "dominant_pct") |>
  filter(dominant_pct > 0.95) |>
  arrange(desc(dominant_pct)) |>
  mutate(dominant_pct = percent(dominant_pct, accuracy = 0.1))

nzv_results |>
  kable(
    col.names = c("Variable", "Dominant Category %"),
    caption = "Near-Zero Variance Variables (>95% in single category)"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

```{r}

#| label: nzv-drop-decisions

vars_to_drop <- c(
  "glimepiride_pioglitazone",
  "metformin_pioglitazone", 
  "metformin_rosiglitazone",
  "glipizide_metformin",
  "tolbutamide",
  "miglitol",
  "tolazamide",
  "chlorpropamide",
  "acarbose",
  "glyburide_metformin",
  "nateglinide"
)

# Document retention decisions
retention_notes <- tibble::tibble(
  variable = c("repaglinide", "max_glu_serum"),
  dominant_pct = c("98.6%", "95.2%"),
  decision = c("Retain — reassess in Part 2",
               "Retain — 'None' is clinically meaningful (test not administered)")
)

retention_notes |>
  kable(
    col.names = c("Variable", "Dominant Category %", "Decision"),
    caption = "Near-Zero Variance Variables Retained with Justification"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

## Readmission rate drift across split groups

The 30-day readmission rate declines slightly from training (7.1%) to validation and holdout (5.6%), 
consistent with temporal trends in hospital readmission rates over the study period (1999–2008). 
This outcome rate heterogeneity across splits suggests model calibration should be evaluated 
alongside discrimination metrics in Part 2, and that threshold selection should be performed 
on the validation set rather than assumed from training data.

```{r}

#| label: outcome-by-split

df |>
  group_by(split_group) |>
  summarise(
    n = n(),
    n_readmitted = sum(readmitted_30day),
    readmission_rate = mean(readmitted_30day)
  ) |>
  mutate(
    split_group = factor(split_group, levels = c("train", "validation", "holdout")),
    readmission_rate = percent(readmission_rate, accuracy = 0.1)
  ) |>
  arrange(split_group) |>
  kable(
    col.names = c("Split Group", "N", "Readmitted <30 days", "Readm Rate"),
    caption = "Readmission Rate by Temporal Split Group"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```