---
title: "Exploratory Data Analysis"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
---

```{r}

#| label: setup
#| include: false

options(java.parameters = "--add-opens=java.base/java.nio=ALL-UNNAMED")
source("/Users/cepickering/Documents/Large Folders/Git/uci_diabetes/reports/part1/R/connections.R")

# Set figure defaults at runtime — overrides project-level YAML
# Note: per-chunk fig size overrides must use inline syntax {r fig.height=X}
# NOT hashpipe #| fig.height: X — Quarto book projects ignore hashpipe overrides
knitr::opts_chunk$set(
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)

library(tableone)
library(kableExtra)
library(naniar)
library(skimr)
library(patchwork)
library(scales)
library(caret)
library(corrplot)
library(vcd)

```

## Overview

This section presents an exploratory analysis of the UCI 130-US Hospitals 
Diabetes dataset following preprocessing and feature engineering in dbt. 
The analytic sample consists of **69,584 encounters** after deduplication 
to one encounter per patient, removal of deceased and hospice discharge 
dispositions, neonatal admissions, and encounters with missing primary diagnosis.
The primary outcome is **30-day hospital readmission** (binary).

## Outcome Distribution and Class Imbalance

Patients not readmitted within 30 days outnumber those readmitted by a ratio of approximately 13:1, with a 30-day readmission rate of 6.7%.

```{r}

#| label: outcome-distribution

outcome_counts <- df |>
  count(readmitted_30day) |>
  mutate(
    label = ifelse(readmitted_30day == 1, "Readmitted <30 days", "Not readmitted"),
    pct = n / sum(n),
    pct_label = paste0(label, "\n", comma(n), " (", percent(pct, accuracy = 0.1), ")")
  )

ggplot(outcome_counts, aes(x = label, y = n, fill = factor(readmitted_30day))) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = paste0(comma(n), "\n(", percent(pct, accuracy = 0.1), ")")),
            vjust = -0.3, size = 4) +
  scale_fill_manual(values = c("#4E9BB9", "#C0392B")) +
  scale_y_continuous(labels = comma, limits = c(0, max(outcome_counts$n) * 1.15)) +
  labs(
    title = "30-Day Readmission Outcome Distribution",
    subtitle = paste0("Positive class rate: ",
                      percent(mean(df$readmitted_30day == 1), accuracy = 0.1)),
    x = NULL,
    y = "Number of Encounters"
  ) +
  theme_classic() +
  theme(
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 12),
        axis.text.x = element_text(size = 13),
        axis.text.y = element_text(size = 10)
  )

```

## Missingness Profile

Missing values for race was likely MCAR — small minority of records with no clear pattern -
so mode imputation to "Caucasian" was the reasonable, quick approach used in dbt feature engineering.
There were only three records with missing gender, which is likely a data entry error and were dropped 
entirely. The remaining missingness is detailed below.

```{r}

#| label: missingness-table

miss_summary <- df |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  mutate(pct_missing = percent(n_missing / nrow(df), accuracy = 0.01)) |>
  arrange(desc(n_missing))

```

```{r}

missingness_decisions <- tibble::tibble(
  variable = c("medical_specialty", "diag_3", "diag_2"),
  mechanism = c(
    "High missingness: likely MNAR (not documented for certain specialties)",
    "Structural: not all admissions have tertiary diagnosis",
    "Structural: not all admissions have secondary diagnosis"
  ),
  decision = c(
    "Retain 'Unknown' as explicit category: already handled in specialty_group",
    "Already handled: NULL maps to 'None' in diag_3_group",
    "Already handled: NULL maps to 'None' in diag_2_group"
  )
) |>
  dplyr::left_join(
    miss_summary |> dplyr::select(variable, pct_missing),
    by = "variable"
  ) |>
  dplyr::select(variable, pct_missing, mechanism, decision)

missingness_decisions |>
  kable(
    col.names = c("Variable", "% Missing", "Mechanism", "Handling Decision"),
    caption = "Missingness Assessment and Handling Decisions"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

## Table 1

This frequency table is stratified by the primary outcome of 30-day readmission and includes all candidate predictor variables. Tests for differences between readmission groups and categories of characteristics are shown as well.

```{r}

#| label: table-one

# Verify column names match exactly
# names(df)  # uncomment to inspect if needed

# Define variable lists — all lowercase per janitor::clean_names()
cat_vars <- c(
  "race", "gender", "age_band", "age_group",
  "admission_type_group", "specialty_group",
  "diag_1_group", "diag_2_group", "diag_3_group",
  "max_glu_serum", "a1cresult",
  "metformin", "repaglinide", "nateglinide", "chlorpropamide",
  "glimepiride", "glipizide", "glyburide", "tolbutamide",
  "pioglitazone", "rosiglitazone", "acarbose", "miglitol",
  "tolazamide", "insulin", "glyburide_metformin",
  "glipizide_metformin", "glimepiride_pioglitazone",
  "metformin_rosiglitazone", "metformin_pioglitazone",
  "change", "diabetes_med",
  "readmitted_raw", "split_group"
)

cont_vars <- c(
  "age_midpoint", "time_in_hospital",
  "num_lab_procedures", "num_procedures", "num_medications",
  "number_outpatient", "number_emergency", "number_inpatient",
  "number_diagnoses", "total_prior_encounters",
  "num_med_changes", "num_meds_active"
)

# Prepare data — factor the strata variable with labels
df_tab <- df |>
  mutate(
    readmitted_30day = factor(
      readmitted_30day,
      levels = c(0, 1),
      labels = c("Not Readmitted", "Readm <30d")
    ),
    # Cast continuous vars explicitly
    across(all_of(cont_vars), as.numeric),
    # Factor ordering for key categoricals
    age_group = factor(age_group, levels = c("young", "middle", "senior")),
    age_band = factor(age_band, levels = c(
      "[0-10)", "[10-20)", "[20-30)", "[30-40)", "[40-50)",
      "[50-60)", "[60-70)", "[70-80)", "[80-90)", "[90-100)"
    )),
    admission_type_group = factor(admission_type_group,
      levels = c("Emergency", "Elective", "Newborn", "Unknown")),
    readmitted_raw = factor(readmitted_raw,
      levels = c("NO", ">30", "<30")),
    split_group = factor(split_group,
      levels = c("train", "validation", "holdout")),
    # Medication columns — logical ordering
    across(all_of(c(
      "metformin", "repaglinide", "nateglinide", "chlorpropamide",
      "glimepiride", "glipizide", "glyburide", "tolbutamide",
      "pioglitazone", "rosiglitazone", "acarbose", "miglitol",
      "tolazamide", "insulin", "glyburide_metformin",
      "glipizide_metformin", "glimepiride_pioglitazone",
      "metformin_rosiglitazone", "metformin_pioglitazone",
      "change", "diabetes_med"
    )), ~ factor(., levels = c("No", "Steady", "Up", "Down"))),
    # Diag groups — alphabetical is fine, Diabetes first makes clinical sense
    diag_1_group = factor(diag_1_group, levels = c(
      "Diabetes", "Circulatory", "Respiratory", "Digestive",
      "Genitourinary", "Injury", "Musculoskeletal", "Neoplasms", "Other"
    )),
    diag_2_group = factor(diag_2_group, levels = c(
      "None", "Diabetes", "Circulatory", "Respiratory", "Digestive",
      "Genitourinary", "Injury", "Musculoskeletal", "Neoplasms", "Other"
    )),
    diag_3_group = factor(diag_3_group, levels = c(
      "None", "Diabetes", "Circulatory", "Respiratory", "Digestive",
      "Genitourinary", "Injury", "Musculoskeletal", "Neoplasms", "Other"
    )),
    max_glu_serum = factor(max_glu_serum,
      levels = c("None", "Norm", ">200", ">300")),
    a1cresult = factor(a1cresult,
      levels = c("None", "Norm", ">7", ">8")),
    race = factor(race, levels = c(
      "Caucasian", "AfricanAmerican", "Hispanic", "Asian", "Other"
    )),
    gender = factor(gender, levels = c("Male", "Female"))
  )

```

```{r}

#| label: table-one-render
#| 
# Create Table 1 using tableone
tab1 <- CreateTableOne(
  vars = c(cat_vars, cont_vars),
  strata = "readmitted_30day",
  data = df_tab,
  factorVars = cat_vars,
  addOverall = TRUE
)

tab1_df <- print(
  tab1,
  showAllLevels = FALSE,
  nonnormal = cont_vars,
  quote = FALSE,
  noSpaces = TRUE,
  printToggle = FALSE
)

# Identify header rows (not indented with 3 spaces)
header_rows <- which(!grepl("^   ", rownames(tab1_df)))

# Restore indentation in row names before passing to kable
rownames(tab1_df) <- ifelse(
  grepl("^   ", rownames(tab1_df)),
  paste0("\u00a0\u00a0\u00a0\u00a0\u00a0\u00a0", trimws(rownames(tab1_df))),
  rownames(tab1_df)
)

tab1_df |>
  kable(
    format = "html",
    escape = FALSE,
    caption = "Table 1: Patient Characteristics Stratified by 30-Day Readmission"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    font_size = 12,
    full_width = TRUE
  ) |>
  row_spec(header_rows, extra_css = "border-top: 2px solid #929393ff;") |>
  scroll_box(height = "800px")

```

## Near-zero variance flagging

Seven variables have zero variance and therefore carry no information. Another six variables 
have >95% of values in a single category, which may cause modeling issues and should be considered 
for removal or re-categorization. The table below summarizes these near-zero variance variables.
Repaglinide and max_glu_serum may be retained for modeling, but the others should not be. This is reassessed later.

```{r}

#| label: near-zero-variance

nzv_results <- df |>
  select(all_of(cat_vars)) |>
  mutate(across(everything(), as.character)) |>
  summarise(across(everything(), ~ {
    tbl <- table(.)
    max(tbl) / sum(tbl)
  })) |>
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "dominant_pct") |>
  filter(dominant_pct > 0.95) |>
  arrange(desc(dominant_pct)) |>
  mutate(dominant_pct = percent(dominant_pct, accuracy = 0.1))

nzv_results |>
  kable(
    col.names = c("Variable", "Dominant Category %"),
    caption = "Near-Zero Variance Variables (>95% in single category)"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

```{r}

#| label: nzv-drop-decisions

vars_to_drop <- c(
  "glimepiride_pioglitazone",
  "metformin_pioglitazone", 
  "metformin_rosiglitazone",
  "glipizide_metformin",
  "tolbutamide",
  "miglitol",
  "tolazamide",
  "chlorpropamide",
  "acarbose",
  "glyburide_metformin",
  "nateglinide"
)

# Document retention decisions
retention_notes <- tibble::tibble(
  variable = c("repaglinide", "max_glu_serum"),
  dominant_pct = c("98.6%", "95.2%"),
  decision = c("Retain: reassess in Part 2",
               "Retain: 'None' is clinically meaningful (test not administered)")
)

retention_notes |>
  kable(
    col.names = c("Variable", "Dominant Category %", "Decision"),
    caption = "Near-Zero Variance Variables Retained with Justification"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

## Readmission rate drift across split groups

The 30-day readmission rate declines slightly from training (7.1%) to validation and holdout (5.6%), 
consistent with temporal trends in hospital readmission rates over the study period (1999–2008). 
This outcome rate heterogeneity across splits suggests model calibration should be evaluated 
alongside discrimination metrics in Part 2, and that threshold selection should be performed 
on the validation set rather than assumed from training data.

```{r}

#| label: outcome-by-split

df |>
  group_by(split_group) |>
  summarise(
    n = n(),
    n_readmitted = sum(readmitted_30day),
    readmission_rate = mean(readmitted_30day)
  ) |>
  mutate(
    split_group = factor(split_group, levels = c("train", "validation", "holdout")),
    readmission_rate = percent(readmission_rate, accuracy = 0.1)
  ) |>
  arrange(split_group) |>
  kable(
    col.names = c("Split Group", "N", "Readmitted <30 days", "Readm Rate"),
    caption = "Readmission Rate by Temporal Split Group"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

## Continuous Variable Distributions

The following is more to evaluate overall distributions of continuous variables and identify any glaring data quality issues than to compare readmission groups, but the plots are colored by readmission status for visual reference.

```{r}

#| label: continuous-distributions
#| fig-height: 10

cont_plot_vars <- c(
  "time_in_hospital", "num_lab_procedures", "num_procedures",
  "num_medications", "number_diagnoses", "total_prior_encounters",
  "num_med_changes", "num_meds_active", "age_midpoint"
)

df_tab |>
  select(all_of(cont_plot_vars), readmitted_30day) |>
  pivot_longer(
    cols = all_of(cont_plot_vars),
    names_to = "variable",
    values_to = "value"
  ) |>
  mutate(variable = str_replace_all(variable, "_", " ") |> str_to_title()) |>
  ggplot(aes(x = value, fill = readmitted_30day)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  scale_fill_manual(values = c("#4E9BB9", "#C0392B")) +
  labs(
    title = "Continuous Variable Distributions by Readmission Status",
    x = NULL,
    y = "Count",
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", 
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))

```

## Categorical Variable Profiles

### Age Distribution

Readmission rates are highest among the oldest age bands, with a notable increase starting around age 60, suggesting age-related factors may be important predictors of readmission risk.

```{r}

#| label: age-distribution

df_tab |>
  count(age_band, readmitted_30day) |>
  group_by(age_band) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = age_band, y = n, fill = readmitted_30day)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = percent(pct, accuracy = 0.1)),
    position = position_dodge(width = 0.9),
    vjust = -0.3, size = 3
  ) +
  scale_fill_manual(values = c("#4E9BB9", "#C0392B")) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Age Band Distribution by Readmission Status",
    x = "Age Band",
    y = "Count",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", 
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))

```

### Primary Diagnosis Group

Some mild variation in readmission rates is observed across primary diagnosis groups, but nothing major; circulatory, diabetes, and injury top the list.

```{r}

#| label: diag1-distribution

df_tab |>
  count(diag_1_group, readmitted_30day) |>
  group_by(diag_1_group) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = reorder(diag_1_group, n), y = n, fill = readmitted_30day)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = percent(pct, accuracy = 0.1)),
    position = position_dodge(width = 0.9),
    hjust = -0.1, size = 3
  ) +
  coord_flip() +
  scale_fill_manual(values = c("#4E9BB9", "#C0392B")) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Primary Diagnosis Group by Readmission Status",
    x = NULL,
    y = "Count",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", 
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))

```

### Admission Type

Admission type distribution is quite similar across readmission groups. A limited number in the newborn category is expected given the exclusion of neonatal admissions, and the "Unknown" category likely reflects data quality issues rather than a clinically meaningful group.

```{r}

#| label: admission-type-distribution

df_tab |>
  count(admission_type_group, readmitted_30day) |>
  group_by(admission_type_group) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = admission_type_group, y = n, fill = readmitted_30day)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = percent(pct, accuracy = 0.1)),
    position = position_dodge(width = 0.9),
    vjust = -0.3, size = 3
  ) +
  scale_fill_manual(values = c("#4E9BB9", "#C0392B")) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Admission Type by Readmission Status",
    x = NULL,
    y = "Count",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", 
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))

```

### Medication Utilization Profile

Insulin is the most commonly changed medication, and the only one with more than 12% of encounters showing an "Up" or "Down" status. Other medications more commonly are increased rather than decreased, at different rates.

```{r}

#| label: medication-profile
#| fig-height: 8

med_vars <- c(
  "metformin", "repaglinide", "nateglinide", "glimepiride",
  "glipizide", "glyburide", "pioglitazone", "rosiglitazone",
  "insulin", "change", "diabetes_med"
)

df_tab |>
  select(all_of(med_vars)) |>
  pivot_longer(everything(), names_to = "medication", values_to = "status") |>
  filter(status != "No") |>
  count(medication, status) |>
  mutate(
    medication = str_replace_all(medication, "_", " ") |> str_to_title(),
    status = factor(status, levels = c("Steady", "Up", "Down", "Yes", "Ch"))
  ) |>
  group_by(medication) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = reorder(medication, n), y = n, fill = status)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = percent(pct, accuracy = 0.1)),
    position = position_dodge(width = 0.9),
    hjust = -0.15, size = 3
  ) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("#6a7891ff", "#d66616ff", "#6c349aff")) +
  labs(
    title = "Medication Utilization Profile",
    subtitle = "Encounters where medication was active (excluding No)",
    x = NULL,
    y = "Number of Encounters",
    fill = "Status"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", 
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 13))

```

### HbA1c and Glucose Test Results

Maximum glucose serum and HbA1c result distributions are similar across readmission groups, with the majority of patients having "None" or "Norm" results. However, a slightly higher proportion of patients readmitted within 30 days have elevated glucose levels, suggesting this may be relevant predictors of readmission risk.

```{r}

#| label: lab-results

p1 <- df_tab |>
  count(a1cresult, readmitted_30day) |>
  group_by(a1cresult) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = a1cresult, y = n, fill = readmitted_30day)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = percent(pct, accuracy = 0.1)),
    position = position_dodge(width = 0.9),
    vjust = -0.3, size = 3
  ) +
  scale_fill_manual(values = c("#4E9BB9", "#C0392B")) +
  scale_y_continuous(labels = comma) +
  labs(title = "HbA1c Result", x = NULL, y = "Count", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", 
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))

p2 <- df_tab |>
  count(max_glu_serum, readmitted_30day) |>
  group_by(max_glu_serum) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = max_glu_serum, y = n, fill = readmitted_30day)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = percent(pct, accuracy = 0.1)),
    position = position_dodge(width = 0.9),
    vjust = -0.3, size = 3
  ) +
  scale_fill_manual(values = c("#4E9BB9", "#C0392B")) +
  scale_y_continuous(labels = comma) +
  labs(title = "Max Glucose Serum", x = NULL, y = "Count", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", 
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))

p1 + p2 + plot_layout(guides = "collect") &
  theme(legend.position = "top")

```

## Correlation Structure

### Continuous Variables — Spearman Correlation

Time in hospital, number inpatient, total prior encounters, and number of medications are the most strongly correlated with 30-day readmission among continuous variables, with Spearman correlation coefficients around 0.1. Other moderate to strong correlations are observed among predictors themselves, such as time in hospital with number of medications and number inpatient, which may have implications for multicollinearity in modeling. 

```{r}

#| label: correlation-heatmap
#| fig-height: 8
#| fig-width: 9

cont_corr_vars <- c(
  "age_midpoint", "time_in_hospital",
  "num_lab_procedures", "num_procedures", "num_medications",
  "number_outpatient", "number_emergency", "number_inpatient",
  "number_diagnoses", "total_prior_encounters",
  "num_med_changes", "num_meds_active", "readmitted_30day"
)

corr_matrix <- df |>
  select(all_of(cont_corr_vars)) |>
  mutate(across(everything(), as.numeric)) |>
  cor(method = "spearman", use = "pairwise.complete.obs")

# Clean up labels
colnames(corr_matrix) <- colnames(corr_matrix) |>
  str_replace_all("_", " ") |>
  str_to_title()
rownames(corr_matrix) <- rownames(corr_matrix)  |>
  str_replace_all("_", " ") |>
  str_to_title()

corrplot(
  corr_matrix,
  method = "color",
  type = "upper",
  order = "hclust",
  tl.cex = 0.8,
  tl.col = "black",
  tl.srt = 45,
  addCoef.col = "black",
  number.cex = 0.65,
  col = colorRampPalette(c("#1d617cff", "white", "#891d11ff"))(200),
  title = "Spearman Correlation, Continuous Variables",
  mar = c(0, 0, 2, 0)
)

```

### Categorical Variables — Cramér's V Association

A few key associations among categorical variables are observed, all logical and validated by literature.

```{r}

#| label: cramers-v-heatmap
#| fig-height: 9
#| fig-width: 9

cat_assoc_vars <- c(
  "race", "gender", "age_group", "admission_type_group",
  "specialty_group", "diag_1_group", "diag_2_group", "diag_3_group",
  "max_glu_serum", "a1cresult", "metformin", "glimepiride",
  "glipizide", "glyburide", "pioglitazone", "insulin",
  "change", "diabetes_med", "readmitted_30day"
)

df_cat <- df |>
  select(all_of(cat_assoc_vars)) |>
  mutate(across(everything(), as.character))

# Compute Cramér's V for all pairs
n_vars <- length(cat_assoc_vars)
v_matrix <- matrix(NA, n_vars, n_vars,
                   dimnames = list(cat_assoc_vars, cat_assoc_vars))

for (i in seq_len(n_vars)) {
  for (j in seq_len(n_vars)) {
    if (i == j) {
      v_matrix[i, j] <- 1
    } else {
      tbl <- table(df_cat[[i]], df_cat[[j]])
      v_matrix[i, j] <- assocstats(tbl)$cramer
    }
  }
}

# Clean labels
clean_labels <- cat_assoc_vars |>
  str_replace_all("_", " ") |>
  str_to_title()
colnames(v_matrix) <- clean_labels
rownames(v_matrix) <- clean_labels

corrplot(
  v_matrix,
  method = "color",
  type = "upper",
  order = "hclust",
  tl.cex = 0.7,
  tl.col = "black",
  tl.srt = 45,
  addCoef.col = "black",
  number.cex = 0.45,
  col = colorRampPalette(c("white", "#891d11ff"))(200),
  is.corr = FALSE,
  title = "Cramer's V Association Matrix, Categorical Variables",
  mar = c(0, 0, 2, 0)
)

```

## Split Group Balance

Despite the slight decline in readmission rates across split groups, demographic and clinical covariates are broadly balanced across train, validation, and holdout sets, supporting the encounter_id-based temporal split as a reasonable proxy for chronological ordering.

```{r}

#| label: split-balance-table

tab2 <- CreateTableOne(
  vars = c(cat_vars, cont_vars),
  strata = "split_group",
  data = df_tab,
  factorVars = cat_vars,
  addOverall = TRUE
)

tab2_df <- print(
  tab2,
  showAllLevels = FALSE,
  nonnormal = cont_vars,
  quote = FALSE,
  noSpaces = TRUE,
  printToggle = FALSE
)

# Identify header rows (not indented with 3 spaces)
header_rows <- which(!grepl("^   ", rownames(tab2_df)))

# Restore indentation in row names before passing to kable
rownames(tab2_df) <- ifelse(
  grepl("^   ", rownames(tab2_df)),
  paste0("\u00a0\u00a0\u00a0\u00a0\u00a0\u00a0", trimws(rownames(tab2_df))),
  rownames(tab2_df)
)

tab2_df |>
  kable(
    format = "html",
    escape = FALSE,
    caption = "Table 2: Patient Characteristics Stratified by Temporal Split Group"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    font_size = 12,
    full_width = TRUE
  ) |>
  row_spec(header_rows, extra_css = "border-top: 2px solid #929393ff;") |>
  scroll_box(height = "800px")

```

## Key Findings and Modeling Implications

```{r}

#| label: key-findings
#| include: false

# Summary statistics for narrative
readmission_rate <- mean(df$readmitted_30day) |> scales::percent(accuracy = 0.1)
n_total <- nrow(df) |> scales::comma()

```

The analytic sample of `r n_total` encounters exhibits a **`r readmission_rate` 
30-day readmission rate**, representing meaningful class imbalance that will 
require explicit handling in Part 2 via class weighting and/or threshold tuning. 
The positive class rate declines modestly from training (7.1%) to validation 
and holdout (5.6%), consistent with temporal trends in hospital readmission 
rates over the study period. Model calibration should be evaluated alongside 
discrimination metrics and threshold selection performed on the validation set.

**Missingness** is largely resolved at the feature engineering stage. Medical 
specialty (48.1% missing) is handled via the `specialty_group` consolidation. 
Race (2.7% missing) will likely receive mode imputation in Part 2. Secondary and 
tertiary diagnosis missingness maps to explicit "None" categories. Three 
records with missing gender will be dropped.

**Eleven near-zero variance medication variables** will be dropped in Part 2 
preprocessing: `glimepiride_pioglitazone`, `metformin_pioglitazone`, 
`metformin_rosiglitazone`, `glipizide_metformin`, `tolbutamide`, `miglitol`, 
`tolazamide`, `chlorpropamide`, `acarbose`, `glyburide_metformin`, and 
`nateglinide`. `repaglinide` is retained provisionally pending further 
assessment. `max_glu_serum` is retained despite 95.2% "None" dominance as 
test presence carries clinical meaning.

**Correlation analysis** reveals expected relationships among utilization 
variables (`number_inpatient`, `number_emergency`, `total_prior_encounters`) 
and moderate associations between medication variables and `diabetes_med`. 
The Cramer's V matrix confirms that diagnosis groups, admission type, and 
insulin status show the strongest associations with readmission among 
categorical variables.

**Split group balance** is broadly maintained across train, validation, and 
holdout sets for demographic and clinical covariates, supporting the 
encounter_id-based temporal split as a reasonable proxy for chronological 
ordering. The outcome rate heterogeneity noted above is the primary 
imbalance of concern and is documented for handling in Part 2.