---
title: "Causal Inference"
execute:
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
---

```{r}

#| label: setup
#| include: false

options(java.parameters = "--add-opens=java.base/java.nio=ALL-UNNAMED")
source("/Users/cepickering/Documents/Large Folders/Git/uci_diabetes/reports/part1/R/connections.R")

# Set figure defaults at runtime — overrides project-level YAML
# Note: per-chunk fig size overrides must use inline syntax {r fig.height=X}
# NOT hashpipe #| fig.height: X — Quarto book projects ignore hashpipe overrides
knitr::opts_chunk$set(
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)

library(tableone)
library(kableExtra)
library(MatchIt)
library(WeightIt)
library(cobalt)
library(broom)
library(patchwork)
library(scales)
library(dagitty)
library(ggdag)

```

## Overview

This section presents three causal inference analyses examining the effect of 
distinct clinical interventions on 30-day hospital readmission. Each analysis 
uses a different identification strategy to demonstrate methodological breadth:

| Analysis | Treatment | Method | Estimand |
|---|---|---|---|
| 1 | HbA1c testing | IPTW with stabilized weights | ATE |
| 2 | Insulin adjustment | Propensity score matching (1:1) | ATT |
| 3 | Any prior inpatient visit | Doubly robust AIPW | ATE |

All analyses use the training split only (`split_group == 'train'`) to avoid 
any information leakage from validation or holdout sets. Sensitivity analyses 
use E-values to assess robustness to unmeasured confounding.

```{r}

#| label: training-data

# Restrict to training split for all causal analyses
df_train <- df |>
  dplyr::filter(split_group == "train")

cat("Training set N:", nrow(df_train), "\n")
cat("Outcome rate:", round(mean(df_train$readmitted_30day), 4), "\n")

```

---

## Analysis 1: Effect of HbA1c Testing on 30-Day Readmission

### Background and Motivation

HbA1c measurement was the central focus of the original Strack et al. (2014) 
analysis of this dataset, which found that testing was associated with reduced 
readmission rates. Here we estimate the Average Treatment Effect (ATE) of 
HbA1c testing using Inverse Probability of Treatment Weighting (IPTW) with 
stabilized weights, adjusting for a richer covariate set than the original paper.

### Causal Assumptions and DAG

```{r}

#| label: dag-hba1c
#| fig-height: 6

dag_hba1c <- dagitty('dag {
  Age -> HbA1c
  Age -> Readm
  Prim_Diag -> HbA1c
  Prim_Diag -> Readm
  Adm_Type -> HbA1c
  Adm_Type -> Readm
  Specialty -> HbA1c
  LabOrdProp -> HbA1c
  Num_Meds -> Readm
  Num_Meds -> HbA1c
  Prior_Inpt -> Readm
  Prior_Inpt -> HbA1c
  Dis_Sev [unobserved]
  Dis_Sev -> HbA1c
  Dis_Sev -> Readm
  HbA1c -> Readm
}')

ggdag(dag_hba1c, layout = "sugiyama",
      node_size = 26,
      text_size = 3) +
  theme_dag() +
  labs(title = "DAG: Effect of HbA1c Testing on Readmission",
       subtitle = "Disease severity is unobserved confounder") +
  theme(legend.position = "none", 
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 12))

```

The key identification assumption is **conditional ignorability** — that 
conditional on observed covariates (age, primary diagnosis, admission type, 
specialty, number of medications, number of lab procedures, and prior 
inpatient visits), HbA1c testing is as-good-as-randomly assigned. The 
unobserved confounder (overall disease severity) represents the primary 
threat to this assumption and can be addressed via sensitivity analysis.

### Treatment Definition and Prevalence

```{r}

#| label: hba1c-treatment

df_train <- df_train |>
  mutate(
    hba1c_tested = as.integer(a1cresult != "None"),
    age_midpoint = as.numeric(age_midpoint),
    time_in_hospital = as.numeric(time_in_hospital),
    num_lab_procedures = as.numeric(num_lab_procedures),
    num_medications = as.numeric(num_medications),
    number_inpatient = as.numeric(number_inpatient),
    number_diagnoses = as.numeric(number_diagnoses),
    total_prior_encounters = as.numeric(total_prior_encounters)
  )

cat("HbA1c tested (treated):", sum(df_train$hba1c_tested), 
    sprintf("(%.1f%%)\n", 100 * mean(df_train$hba1c_tested)))
cat("HbA1c not tested (control):", sum(df_train$hba1c_tested == 0),
    sprintf("(%.1f%%)\n", 100 * mean(df_train$hba1c_tested == 0)))

```

### Propensity Score Estimation

```{r}

#| label: hba1c-ps

# Propensity score model
ps_formula_hba1c <- hba1c_tested ~ 
  age_midpoint + 
  diag_1_group + 
  admission_type_group + 
  specialty_group +
  num_medications +
  num_lab_procedures +
  number_inpatient +
  number_diagnoses +
  race +
  gender

ps_model_hba1c <- glm(
  ps_formula_hba1c,
  data = df_train,
  family = binomial()
)

df_train <- df_train |>
  mutate(ps_hba1c = predict(ps_model_hba1c, type = "response"))

# Check overlap
ggplot(df_train, aes(x = ps_hba1c, fill = factor(hba1c_tested))) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(
    values = c("#4E9BB9", "#C0392B"),
    labels = c("Not Tested", "Tested")
  ) +
  labs(
    title = "Propensity Score Overlap — HbA1c Testing",
    x = "Estimated Propensity Score",
    y = "Density",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top",  
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))

```

### IPTW Weights and Balance Assessment

```{r}

#| label: hba1c-iptw

# Stabilized IPTW weights
w_hba1c <- weightit(
  ps_formula_hba1c,
  data = df_train,
  method = "ps",
  estimand = "ATE",
  stabilize = TRUE
)

summary(w_hba1c)

# Balance assessment using cobalt
bal_hba1c <- bal.tab(
  w_hba1c,
  stats = c("m", "ks"),
  thresholds = c(m = 0.1),
  un = TRUE
)

love.plot(
  bal_hba1c,
  threshold = 0.1,
  abs = TRUE,
  colors = c("#C0392B", "#4E9BB9"),
  shapes = c("circle", "triangle"),
  size = 3,
  title = "Covariate Balance Before and After IPTW — HbA1c Testing"
)

```

### Effect Estimation

```{r}

#| label: hba1c-effect

# Weighted outcome model
outcome_model_hba1c <- glm(
  readmitted_30day ~ hba1c_tested,
  data = df_train,
  weights = w_hba1c$weights,
  family = quasibinomial()
)

tidy(outcome_model_hba1c, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term == "hba1c_tested") |>
  mutate(across(where(is.numeric), ~ round(., 3))) |>
  kable(
    col.names = c("Term", "OR", "Std Error", "Statistic", 
                  "P-value", "L95", "U95"),
    caption = "IPTW Estimated Effect of HbA1c Testing on 30-Day Readmission"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

The odds ratio of `r round(exp(coef(outcome_model_hba1c)["hba1c_tested"]), 3)` suggests that patients who were tested for HbA1c had approximately 8% lower odds of 30-day readmission compared to those who were not tested, after adjusting for confounders via IPTW. The statistical signifiacnce is suggestive, but does not exceed an alpha of 0.05. This is directionally consistent with the original Strack et al. (2014) finding of a protective association between HbA1c testing and readmission; this analysis, however, used a richer covariate set and a different estimation method (IPTW vs multivariable regression).

### Sensitivity Analysis — E-value

```{r}

#| label: hba1c-evalue

# E-value calculation
# E-value = OR + sqrt(OR * (OR - 1)) for OR > 1
# For OR < 1, use 1/OR first
or_hba1c <- exp(coef(outcome_model_hba1c)["hba1c_tested"])

or_for_evalue <- ifelse(or_hba1c < 1, 1/or_hba1c, or_hba1c)
evalue_hba1c <- or_for_evalue + sqrt(or_for_evalue * (or_for_evalue - 1))

cat("E-value:", round(evalue_hba1c, 3), "\n")

```

The E-value quantifies the minimum strength of association that an unmeasured confounder would need to have with both HbA1c testing and 30-day readmission to fully explain away the observed association. The E-value of `r round(evalue_hba1c, 2)` suggests that the observed effect is fairly robust to potential unmeasured confounding.

---

## Analysis 2: Effect of Insulin Adjustment on 30-Day Readmission

### Background and Motivation

Insulin management is a central component of inpatient diabetes care. We 
estimate the Average Treatment Effect on the Treated (ATT) of active insulin 
dose adjustment (increase or decrease) versus no active change (no insulin or 
stable dose) using 1:1 nearest-neighbor propensity score matching with a 
caliper.

### Treatment Definition
```{r}

#| label: insulin-treatment

df_train <- df_train |>
  mutate(
    insulin_adjusted = as.integer(insulin %in% c("Up", "Down")),
    max_glu_num = case_when(
      max_glu_serum == "None" ~ 0,
      max_glu_serum == "Norm" ~ 1,
      max_glu_serum == ">200" ~ 2,
      max_glu_serum == ">300" ~ 3
    ),
    a1c_num = case_when(
      a1cresult == "None" ~ 0,
      a1cresult == "Norm" ~ 1,
      a1cresult == ">7"   ~ 2,
      a1cresult == ">8"   ~ 3
    )
  )

cat("Insulin adjusted (treated):", sum(df_train$insulin_adjusted),
    sprintf("(%.1f%%)\n", 100 * mean(df_train$insulin_adjusted)))
cat("No active change (control):", sum(df_train$insulin_adjusted == 0),
    sprintf("(%.1f%%)\n", 100 * mean(df_train$insulin_adjusted == 0)))

```

### DAG
```{r}

#| label: dag-insulin

dag_insulin <- dagitty('dag {
  Age -> Insulin_Adj
  Age -> Readm
  PrimDiag -> Insulin_Adj
  PrimDiag -> Readm
  HbA1c -> Insulin_Adj
  Glucose -> Insulin_Adj
  Num_Meds -> Insulin_Adj
  Num_Meds -> Readm
  TimeHosp -> Insulin_Adj
  Adm_Type -> Insulin_Adj
  Adm_Type -> Readm
  Num_Diag -> Readm
  Num_Diag -> Insulin_Adj
  Dis_Sev [unobserved]
  Dis_Sev -> Insulin_Adj
  Dis_Sev -> Readm
  Insulin_Adj -> Readm
}')

ggdag(dag_insulin, layout = "sugiyama",
      node_size = 26,
      text_size = 3) +
  theme_dag() +
  labs(title = "DAG: Effect of Insulin Adjustment on Readmission",
       subtitle = "Disease severity is an unobserved confounder") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 12))

```

### Propensity Score Matching
```{r}

#| label: insulin-psm

ps_formula_insulin <- insulin_adjusted ~
  age_midpoint +
  diag_1_group +
  admission_type_group +
  a1c_num +
  max_glu_num +
  num_medications +
  time_in_hospital +
  number_diagnoses +
  race +
  gender

match_insulin <- matchit(
  ps_formula_insulin,
  data = df_train,
  method = "nearest",
  distance = "glm",
  ratio = 1,
  caliper = 0.1,
  replace = FALSE
)

summary(match_insulin, un = FALSE)

```

### Balance Assessment
```{r}

#| label: insulin-balance

love.plot(
  match_insulin,
  threshold = 0.1,
  abs = TRUE,
  colors = c("#C0392B", "#4E9BB9"),
  shapes = c("circle", "triangle"),
  size = 3,
  title = "Covariate Balance Before and After PSM — Insulin Adjustment"
)

```

### Effect Estimation
```{r}

#| label: insulin-effect

matched_data_insulin <- match.data(match_insulin)

outcome_model_insulin <- glm(
  readmitted_30day ~ insulin_adjusted,
  data = matched_data_insulin,
  weights = weights,
  family = quasibinomial()
)

tidy(outcome_model_insulin, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term == "insulin_adjusted") |>
  mutate(across(where(is.numeric), ~ round(., 3))) |>
  kable(
    col.names = c("Term", "OR", "Std Error", "Statistic",
                  "P-value", "L95", "U95"),
    caption = "PSM Estimated Effect of Insulin Adjustment on 30-Day Readmission"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

The odds ratio of `r round(exp(coef(outcome_model_insulin)["insulin_adjusted"]), 3)` suggests that patients with active insulin adjustment had approximately 17% higher odds of 30-day readmission compared to those with no active change, after adjusting for confounders via PSM. This was statistically significant at the 0.05 level. 

### Sensitivity Analysis — E-value

```{r}
#| label: insulin-evalue

or_insulin <- exp(coef(outcome_model_insulin)["insulin_adjusted"])
or_for_evalue_ins <- ifelse(or_insulin < 1, 1/or_insulin, or_insulin)
evalue_insulin <- or_for_evalue_ins + sqrt(or_for_evalue_ins * (or_for_evalue_ins - 1))

cat("E-value:", round(evalue_insulin, 3), "\n")

```

The E-value of `r round(evalue_insulin, 2)` suggests that the observed effect is quite robust to potential unmeasured confounding,
as an unmeasured confounder would need to be associated with both treatment and outcome by a fairly high factor.

---

## Analysis 3: Effect of Prior Inpatient Visits on 30-Day Readmission

### Background and Motivation

Prior inpatient utilization is among the strongest known predictors of 
hospital readmission. We estimate the ATE of having had any prior inpatient 
visit in the year before the index admission using doubly robust Augmented 
Inverse Probability Weighting (AIPW), which is consistent if either the 
propensity score model or the outcome model is correctly specified. 
This method is used here because of the strong confounding by chronic disease burden, 
which is only partially captured by observed covariates. 
AIPW provides some protection against model misspecification and 
allows for flexible machine learning models in both components if desired.

### Treatment Definition

```{r}

#| label: prior-inpatient-treatment

df_train <- df_train |>
  mutate(prior_inpatient = as.integer(number_inpatient > 0))

cat("Any prior inpatient (treated):", sum(df_train$prior_inpatient),
    sprintf("(%.1f%%)\n", 100 * mean(df_train$prior_inpatient)))
cat("No prior inpatient (control):", sum(df_train$prior_inpatient == 0),
    sprintf("(%.1f%%)\n", 100 * mean(df_train$prior_inpatient == 0)))

```

### DAG

```{r}

#| label: dag-prior-inpatient

dag_prior <- dagitty('dag {
  Age -> Prior_Inp
  Age -> Readm
  Prim_Diag -> Prior_Inp
  Prim_Diag -> Readm
  Num_Diag -> Prior_Inp
  Num_Diag -> Readm
  Adm_Type -> Prior_Inp
  Adm_Type -> Readm
  NumMeds -> Prior_Inp
  NumMeds -> Readm
  Race -> Prior_Inp
  Race -> Readm
  ChrDis_Bur -> Prior_Inp
  ChrDis_Bur -> Readm
  Prior_Inp -> Readm
}')

ggdag(dag_prior, layout = "sugiyama",
      node_size = 26,
      text_size = 3) +
  theme_dag() +
  labs(title = "DAG: Effect of Prior Inpatient Visit on Readmission",
       subtitle = "Chronic disease burden is an unobserved confounder") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 12))

```

### Doubly Robust AIPW Estimation

```{r}

#| label: prior-inpatient-aipw

ps_formula_prior <- prior_inpatient ~
  age_midpoint +
  diag_1_group +
  number_diagnoses +
  num_medications +
  admission_type_group +
  race +
  gender +
  time_in_hospital

w_prior <- weightit(
  ps_formula_prior,
  data = df_train,
  method = "ps",
  estimand = "ATE",
  stabilize = TRUE
)

# Balance check
love.plot(
  w_prior,
  threshold = 0.1,
  abs = TRUE,
  colors = c("#C0392B", "#4E9BB9"),
  shapes = c("circle", "triangle"),
  size = 3,
  title = "Covariate Balance Before and After IPTW — Prior Inpatient Visit"
)

```

```{r}

#| label: aipw-estimation

# Doubly robust AIPW via outcome model on weighted data
# Propensity score component
ps_prior <- predict(
  glm(ps_formula_prior, data = df_train, family = binomial()),
  type = "response"
)

# Outcome model components
outcome_formula_prior <- readmitted_30day ~
  prior_inpatient +
  age_midpoint +
  diag_1_group +
  number_diagnoses +
  num_medications +
  admission_type_group +
  race +
  gender +
  time_in_hospital

# Fit outcome model under treatment = 1 and treatment = 0
df_t1 <- df_train |> mutate(prior_inpatient = 1)
df_t0 <- df_train |> mutate(prior_inpatient = 0)

outcome_model_prior <- glm(
  outcome_formula_prior,
  data = df_train,
  family = binomial()
)

mu1 <- predict(outcome_model_prior, newdata = df_t1, type = "response")
mu0 <- predict(outcome_model_prior, newdata = df_t0, type = "response")

# AIPW estimator
A <- df_train$prior_inpatient
Y <- df_train$readmitted_30day
ps <- ps_prior

aipw_1 <- mean(mu1 + (A / ps) * (Y - mu1))
aipw_0 <- mean(mu0 + ((1 - A) / (1 - ps)) * (Y - mu0))
aipw_ate <- aipw_1 - aipw_0

# Bootstrap SE
set.seed(49563)
n_boot <- 300 
boot_ates <- replicate(n_boot, {
  idx <- sample(nrow(df_train), replace = TRUE)
  df_b <- df_train[idx, ]
  ps_b <- predict(glm(ps_formula_prior, data = df_b, family = binomial()),
                  type = "response")
  om_b <- glm(outcome_formula_prior, data = df_b, family = binomial())
  mu1_b <- predict(om_b, newdata = df_b |> mutate(prior_inpatient = 1),
                   type = "response")
  mu0_b <- predict(om_b, newdata = df_b |> mutate(prior_inpatient = 0),
                   type = "response")
  A_b <- df_b$prior_inpatient
  Y_b <- df_b$readmitted_30day
  mean(mu1_b + (A_b/ps_b)*(Y_b - mu1_b)) -
    mean(mu0_b + ((1-A_b)/(1-ps_b))*(Y_b - mu0_b))
})

aipw_se <- sd(boot_ates)
aipw_ci <- aipw_ate + c(-1, 1) * 1.96 * aipw_se

tibble::tibble(
  Estimand = "ATE",
  Estimate = round(aipw_ate, 4),
  SE = round(aipw_se, 4),
  CI_Lower = round(aipw_ci[1], 4),
  CI_Upper = round(aipw_ci[2], 4)
) |>
  kable(
    caption = "AIPW Estimated Effect of Prior Inpatient Visit on 30-Day Readmission (Risk Difference)"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

The risk difference of `r round(aipw_ate, 4)` means patients with any prior inpatient visit have a 5.75 percentage point higher absolute probability of 30-day readmission compared to patients with no prior inpatient visit, after adjusting for confounders. Given the standard error estimate, this is statistically significant at the 0.05 level.

### Sensitivity Analysis — E-value

```{r}

#| label: prior-evalue

# Convert risk difference to approximate RR for E-value
baseline_risk <- mean(df_train$readmitted_30day[df_train$prior_inpatient == 0])
rr_prior <- aipw_1 / aipw_0 # was (baseline_risk + aipw_ate) / baseline_risk

rr_for_evalue <- ifelse(rr_prior < 1, 1/rr_prior, rr_prior)
evalue_prior <- rr_for_evalue + sqrt(rr_for_evalue * (rr_for_evalue - 1))

cat("Estimated RR (approximated from AIPW ratio):", round(rr_prior, 3), "\n")
cat("E-value:", round(evalue_prior, 3), "\n")

```

The E-value of `r round(evalue_prior, 2)` suggests that the observed effect is very robust to potential unmeasured confounding. A risk ratio of ~2 means patients with prior inpatient visits are approximately twice as likely to be readmitted within 30 days compared to patients without, after adjusting for potential confounders.

This is clinically reasonable — prior inpatient utilization is one of the strongest known predictors of readmission in the literature. A doubling of risk after adjustment is directionally consistent with other published studies. The fact that this persists after adjusting for the observed confounders suggests either a genuine causal effect or residual confounding from unmeasured chronic disease burden — which is exactly what the E-value is designed to probe.  

---

## Summary of Causal Findings

Here, we summarize key methods and results for all three analyses in a single table for easy comparison. Note the different estimands and metrics.

```{r}

#| label: causal-summary-table

tibble::tibble(
  Analysis = c(
    "HbA1c Testing",
    "Insulin Adjustment",
    "Prior Inpatient Visit"
  ),
  Treatment = c(
    "Tested vs not",
    "Up/down vs no/steady",
    "Any vs none"
  ),
  Estimand = c("ATE", "ATT", "ATE"),
  Method = c("IPTW Stabilized", "PSM 1:1 Nearest Neighbor", "Doubly Robust AIPW"),
  Metric = c("Odds Ratio", "Odds Ratio", "Risk Difference"),
  Estimate = c(
        round(exp(coef(outcome_model_hba1c)["hba1c_tested"]), 3),
        round(exp(coef(outcome_model_insulin)["insulin_adjusted"]), 3),
        round(aipw_ate, 3)
  ),
) |>
  kable(caption = "Summary of Causal Inference Analyses") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)

```