---
title: "Experimental Design"
execute:
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
---

```{r}

#| label: setup
#| include: false

options(java.parameters = "--add-opens=java.base/java.nio=ALL-UNNAMED")
source("/Users/cepickering/Documents/Large Folders/Git/uci_diabetes/reports/part1/R/connections.R")

# Set figure defaults at runtime — overrides project-level YAML
# Note: per-chunk fig size overrides must use inline syntax {r fig.height=X}
# NOT hashpipe #| fig.height: X — Quarto book projects ignore hashpipe overrides
knitr::opts_chunk$set(
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)

library(kableExtra)
library(patchwork)
library(scales)
library(pwr)
library(gsDesign)

```

## Overview

This document designs a prospective randomized controlled trial (RCT) to evaluate 
whether deploying the readmission risk prediction model developed in Part 2 
as a clinical decision support tool reduces 30-day hospital readmission 
rates among diabetic inpatients. The experimental design draws directly on 
empirical estimates from the EDA and causal inference analyses to ground 
power calculations and stopping rules in realistic assumptions.

---

## Section 1: Framing the Experiment

### Clinical Deployment Scenario

The readmission risk model developed in Part 2 generates a predicted 
probability of 30-day readmission for each diabetic inpatient at the time 
of discharge planning. In clinical deployment, this score would be surfaced 
to the care team via the electronic health record (EHR) at discharge. 

However, a risk score alone does not reduce readmissions — the mechanism 
of action requires an **actionable intervention bundle** triggered by the 
score. Patients whose predicted probability exceeds a **pre-specified threshold** 
$\tau$ are flagged as high-risk and receive the following interventions:

```{r}

#| label: intervention-bundle

tibble::tibble(
  Intervention = c(
    "Enhanced discharge planning",
    "Expedited follow-up appointment",
    "Transitional care program enrollment",
    "Pharmacist medication reconciliation",
    "Targeted care gap closure"
  ),
  Description = c(
    "Social work consultation, patient education, and post-discharge support coordination",
    "7-day post-discharge visit scheduled prior to discharge (vs. standard 30-day)",
    "Nurse navigator assigned; structured post-discharge phone calls at days 3, 7, and 14",
    "Full medication review and reconciliation before discharge order is signed",
    "Flagging unordered HbA1c, uninitiated insulin adjustment, or unaddressed comorbidities"
  )
) |>
  kable(caption = "Intervention Bundle Triggered by High-Risk Flag") |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = TRUE
  )

```

This deployment model is consistent with established clinical decision support 
frameworks such as the LACE index and Epic's readmission risk score, both of 
which trigger care coordinator workflows for flagged patients.

### Estimand and Trial Design

The experiment is not a naive full-population RCT. Because the intervention 
is only applied to patients flagged above threshold $\tau$, the trial is an 
**enriched population RCT** — randomization occurs within the flagged 
sub-population only. Patients below threshold receive standard care regardless 
of treatment arm assignment.

The primary estimand is the **Average Treatment Effect among the Treated (ATT)** — 
the effect of the intervention bundle among patients who would be 
flagged by the model:

$$ATT = E[Y(1) - Y(0) \mid \hat{p} \geq \tau]$$

Where:

- $Y(1)$ = potential outcome under intervention bundle

- $Y(0)$ = potential outcome under standard care  

- $\hat{p}$ = predicted readmission probability from Part 2 model

- $\tau$ = decision threshold (to be specified)

### Threshold Selection and Its Implications

The choice of $\tau$ is a critical design parameter that directly determines:

- The **prevalence of high-risk flags** — what proportion of patients are enrolled

- The **baseline readmission rate** in the trial population — higher $\tau$ means higher-risk enrolled patients

- The **statistical power** — higher baseline risk means larger absolute effect sizes for a given relative reduction

```{r}

#| label: threshold-implications

# Use holdout set to simulate threshold-based enrollment
df_holdout <- df |>
  dplyr::filter(split_group == "holdout") |>
  mutate(
    # Simulate predicted probabilities using empirical readmission rate
    # In practice these would come from the Part 2 model
    # Here we use a logistic model on key predictors as a placeholder
    readmitted_30day = as.numeric(readmitted_30day),
    number_inpatient = as.numeric(number_inpatient),
    num_medications = as.numeric(num_medications),
    number_diagnoses = as.numeric(number_diagnoses),
    time_in_hospital = as.numeric(time_in_hospital),
    age_midpoint = as.numeric(age_midpoint)
  )

# Fit placeholder risk model on training data
df_train_exp <- df |>
  dplyr::filter(split_group == "train") |>
  mutate(across(c(readmitted_30day, number_inpatient, num_medications,
                  number_diagnoses, time_in_hospital, age_midpoint), as.numeric))

placeholder_model <- glm(
  readmitted_30day ~ number_inpatient + num_medications + 
    number_diagnoses + time_in_hospital + age_midpoint +
    diag_1_group + admission_type_group,
  data = df_train_exp,
  family = binomial()
)

df_holdout <- df_holdout |>
  mutate(pred_prob = predict(placeholder_model, 
                             newdata = df_holdout, 
                             type = "response"))

# Compute enrollment and baseline risk across thresholds
thresholds <- seq(0.05, 0.25, by = 0.01)

threshold_summary <- purrr::map_dfr(thresholds, function(tau) {
  flagged <- df_holdout |> dplyr::filter(pred_prob >= tau)
  tibble::tibble(
    threshold = tau,
    n_flagged = nrow(flagged),
    pct_flagged = nrow(flagged) / nrow(df_holdout),
    baseline_risk = mean(flagged$readmitted_30day)
  )
})

```

```{r}

# Plot
p1 <- ggplot(threshold_summary, 
             aes(x = threshold, y = pct_flagged)) +
  geom_line(color = "#4E9BB9", linewidth = 1) +
  geom_point(color = "#4E9BB9", size = 2) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(labels = percent) +
  labs(
    title = "% Patients Flagged by Threshold",
    x = "Risk Threshold (τ)",
    y = "% Flagged"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

p2 <- ggplot(threshold_summary,
             aes(x = threshold, y = baseline_risk)) +
  geom_line(color = "#C0392B", linewidth = 1) +
  geom_point(color = "#C0392B", size = 2) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(labels = percent) +
  labs(
    title = "Baseline Readmission Risk by Threshold",
    x = "Risk Threshold (τ)",
    y = "Baseline Readmission Rate"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

p1 + p2

```

```{r}

#| label: threshold-table

threshold_summary |>
  # dplyr::filter(threshold %in% c(0.05, 0.08, 0.1, 0.12, 0.15, 0.2)) |>
  mutate(
    threshold = percent(threshold, accuracy = 1),
    pct_flagged = percent(pct_flagged, accuracy = 0.1),
    baseline_risk = percent(baseline_risk, accuracy = 0.1),
    n_flagged = comma(n_flagged)
  ) |>
  kable(
    col.names = c("Threshold (τ)", "N Flagged", "% Flagged", 
                  "Baseline Readmission Rate"),
    caption = "Enrollment and Baseline Risk by Decision Threshold (Holdout Set)"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )

```

### Selected Design Parameters

Based on the threshold analysis above, we select $\tau = 0.1$ as the 
primary design threshold, which balances:

- **Clinical feasibility** — flagging a manageable proportion of patients 
  for enhanced intervention
- **Statistical efficiency** — sufficient baseline risk to detect clinically 
  meaningful absolute risk reductions
- **Operational realism** — consistent with deployment thresholds used in 
  published clinical decision support literature

Predicted probabilities are generated from a placeholder logistic regression model for illustrative purposes. The threshold analysis will be updated with calibrated predictions from the Part 2 XGBoost and neural network models, which are expected to produce better-discriminating risk scores and more realistic flagging rates.

```{r}

#| label: design-parameters

# Extract parameters at selected threshold
selected <- threshold_summary |> 
  dplyr::filter(abs(threshold - 0.1) < 0.001)

cat("Selected threshold (τ):", "10%\n")
cat("Estimated % patients flagged:", 
    percent(selected$pct_flagged, accuracy = 0.1), "\n")
cat("Estimated baseline readmission rate (flagged):", 
    percent(selected$baseline_risk, accuracy = 0.1), "\n")
cat("Estimated N flagged per year (assuming ~5,000 annual inpatient admissions):",
    round(0.067 * selected$pct_flagged * 5000), "\n")

```

The baseline readmission rate among flagged patients at $\tau = 0.1$ 
directly informs the power calculations in Section 2, replacing the 
full-cohort rate of ~6.7% with the enriched population rate — a 
meaningfully different and more realistic assumption for trial planning.

## Section 2: Power Analysis

### Minimum Detectable Effect and Clinical Meaningfulness

The primary power analysis targets the **absolute risk reduction (ARR)** 
in 30-day readmission rate among flagged patients receiving the intervention 
bundle versus standard care. We define the **minimum detectable effect (MDE)** 
as the smallest ARR that would be considered clinically meaningful given 
operational costs of the intervention.

Published transitional care interventions for high-risk diabetic patients 
report ARRs ranging from 3-8 percentage points. Given our baseline readmission 
rate of 11.2% among flagged patients at τ=0.1, we consider:

```{r}

#| label: mde-table

tibble::tibble(
  ARR = c(0.02, 0.03, 0.04, 0.05, 0.06),
  RRR = c(0.02, 0.03, 0.04, 0.05, 0.06) / selected$baseline_risk,
  p_control = selected$baseline_risk,
  p_treatment = selected$baseline_risk - c(0.02, 0.03, 0.04, 0.05, 0.06)
) |>
  mutate(
    ARR = percent(ARR, accuracy = 0.1),
    RRR = percent(RRR, accuracy = 0.1),
    p_control = percent(p_control, accuracy = 0.1),
    p_treatment = percent(p_treatment, accuracy = 0.1)
  ) |>
  kable(
    col.names = c("Abs Risk Reduction", "Rel Risk Reduction", 
                  "Ctrl Rate", "Tx Rate"),
    caption = paste0("Effect Size Scenarios at Baseline Risk of ", round(selected$baseline_risk, 3)*100, "%")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )

```

A **3 percentage point ARR** (26.7% relative reduction) is selected as the 
primary MDE — conservative relative to published literature but operationally 
meaningful given the cost of the intervention bundle.

### Sample Size Requirements

```{r}

#| label: sample-size

# Parameters
p_control <- selected$baseline_risk
alpha_levels <- c(0.05, 0.01)
power_levels <- c(0.8, 0.9)
arr_values <- c(0.02, 0.03, 0.04, 0.05, 0.06)

# Compute sample size per arm for all combinations
ss_results <- purrr::map_dfr(alpha_levels, function(alpha) {
  purrr::map_dfr(power_levels, function(pwr_level) {
    purrr::map_dfr(arr_values, function(arr) {
      p_treat <- p_control - arr
      test <- pwr.2p.test(
        h = ES.h(p_control, p_treat),
        sig.level = alpha,
        power = pwr_level,
        alternative = "two.sided"
      )
      tibble::tibble(
        alpha = alpha,
        power = pwr_level,
        arr = arr,
        n_per_arm = ceiling(test$n),
        n_total = ceiling(test$n) * 2
      )
    })
  })
})

ss_results |>
  mutate(
    arr = percent(arr, accuracy = 1),
    alpha = as.character(alpha),
    power = percent(power, accuracy = 1),
    n_per_arm = comma(n_per_arm),
    n_total = comma(n_total)
  ) |>
  kable(
    col.names = c("Alpha", "Power", "ARR", 
                  "N per Arm", "Total N"),
    caption = "Sample Size Requirements by Effect Size, Alpha, and Power"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) |>
  collapse_rows(columns = 1:2, valign = "top")

```

### Power Curves

```{r, fig.height=9}

#| label: power-curves

# Power curves across sample sizes for primary scenarios
n_range <- seq(100, 3000, by = 50)

power_curve_data <- purrr::map_dfr(arr_values, function(arr) {
  p_treat <- p_control - arr
  h <- ES.h(p_control, p_treat)
  purrr::map_dfr(n_range, function(n) {
    pwr_result <- pwr.2p.test(
      h = h,
      n = n,
      sig.level = 0.05,
      alternative = "two.sided"
    )
    tibble::tibble(
      arr = paste0("ARR = ", percent(arr, accuracy = 1)),
      n_per_arm = n,
      power = pwr_result$power
    )
  })
})

ggplot(power_curve_data, 
       aes(x = n_per_arm, y = power, color = arr)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", 
             color = "gray50", linewidth = 0.9) +
  geom_hline(yintercept = 0.90, linetype = "dotted",
             color = "gray50", linewidth = 0.9) +
  annotate("text", x = 3000, y = 0.82, 
           label = "80% power", hjust = 1, size = 4) +
  annotate("text", x = 3000, y = 0.92,
           label = "90% power", hjust = 1, size = 4) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  scale_x_continuous(labels = comma) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Power Curves by Effect Size (α = 0.05)",
    subtitle = "Baseline readmission rate = 11.2% among flagged patients (τ = 0.1)",
    x = "N per arm",
    y = "Statistical power",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top",
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 13))

```

### Feasibility Assessment

```{r}

#| label: feasibility

# Feasibility based on hypothetical hospital volume
# Assumption: medium-sized academic medical center with ~5,000 
# diabetic inpatient admissions per year
# At tau=0.1 flagging rate of 15.7%, implies ~784 flagged per year
annual_admissions <- 5000
flagging_rate <- selected$pct_flagged
annual_flagged <- round(annual_admissions * flagging_rate)
annual_per_arm <- floor(annual_flagged / 2)

# Primary scenario: ARR=3pp, alpha=0.05, power=80%
primary_ss <- ss_results |>
  dplyr::filter(
    arr == 0.03,
    alpha == 0.05,
    power == 0.8
  )

years_required <- round(primary_ss$n_per_arm / annual_per_arm, 1)

tibble::tibble(
  Parameter = c(
    "Selected threshold (τ)",
    "Baseline readmission rate (flagged)",
    "Primary MDE (ARR)",
    "Alpha",
    "Power",
    "Required N per arm",
    "Required total N",
    "Estimated annual flagged patients",
    "Estimated annual per arm",
    "Estimated enrollment duration"
  ),
  Value = c(
    "10%",
    "11.2%",
    "3 perc pts",
    "0.05",
    "80%",
    comma(primary_ss$n_per_arm),
    comma(primary_ss$n_total),
    comma(annual_flagged),
    comma(annual_per_arm),
    paste0(years_required, " years")
  )
) |>
  kable(
    caption = "Primary Trial Design Parameters and Feasibility Summary"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )

```

The feasibility table contextualizes the sample size requirements against 
realistic enrollment rates. If the primary MDE of 3 percentage points requires 
more than about 4 years of enrollment at a single site, a multi-site trial or a 
larger MDE should be considered. The power curves allow stakeholders to 
explicitly trade off effect size, sample size, and enrollment duration when 
making trial design decisions.

Above, we see that a medium-sized academic medical center with around 5,000 diabetic inpatient 
admissions annually, the primary trial scenario (ARR = 3pp, alpha = 0.05, 
80% power) requires approximately 4 years of enrollment at a single site. 
This duration is at the upper bound of operational feasibility for a 
single-site RCT. Three practical alternatives merit consideration:

1. Multi-site trial: Enrolling across 3-4 hospital systems would reduce 
   enrollment to approximately 12-16 months — strongly preferred.

2. Relaxed MDE: Accepting a 4pp ARR as the primary MDE reduces required 
   N per arm to 813, cutting enrollment to approximately 2 years at a 
   single site.

3. Adaptive design: A group sequential design with interim analyses 
   (Section 3) allows early stopping for efficacy if the effect is larger 
   than the MDE, potentially reducing realized enrollment duration.

The power curves illustrate that moving from ARR=3pp to ARR=4pp 
meaningfully reduces sample size while remaining within the range of 
effects reported in published transitional care literature.

## Section 3: Group Sequential Design with Interim Analyses

### Motivation

A fixed-sample RCT requiring 4 years of enrollment at a single site carries 
substantial operational risk — if the true effect is larger than the MDE, 
the trial runs longer than necessary. A **group sequential design** addresses 
this by pre-specifying interim analyses at which the trial may stop early for 
efficacy or futility, while controlling the overall Type I error rate.

### Design Parameters

We specify a group sequential design with:

- **4 analyses**: 3 interim analyses and 1 final analysis, 
  corresponding to one interim analysis per year of enrollment

- **O'Brien-Fleming spending function**: conservative early stopping boundaries 
  that preserve most alpha for the final analysis

- **Primary scenario**: ARR = 3pp, baseline = 11.2%, alpha = 0.05, power = 80%

- **Two-sided test** with symmetric boundaries

```{r}

#| label: gs-design

# Group sequential design
gs <- gsDesign(
  k = 4,                    # 4 analyses (3 interim + 1 final)
  test.type = 2,            # two-sided symmetric
  alpha = 0.05,
  beta = 0.20,              # 1 - power = 0.20 for 80% power
  sfu = sfLDOF,             # O'Brien-Fleming spending function
  delta1 = 0.03 / 0.112,   # effect size as proportion of control rate
  n.fix = 1524 * 2          # fixed sample size from power analysis
)

# Display design summary
print(gs)

```

### Stopping Boundaries

```{r fig.height=7}

#| label: gs-boundaries

plot(gs, 
     plottype = 1,
     main = "O'Brien-Fleming Stopping Boundaries",
     xlab = "Analysis",
     ylab = "Z Statistic")

```

### Sample Size and Timing

```{r}

#| label: gs-timing

# Extract design parameters
timing <- gs$timing
n_analyses <- gs$n.I

tibble::tibble(
  Analysis = c("Interim 1", "Interim 2", "Interim 3", "Final"),
  Information_Fraction = round(timing, 3),
  N_per_arm = ceiling(n_analyses / 2),
  N_total = ceiling(n_analyses),
  Efficacy_Z = round(gs$upper$bound, 3),
  Futility_Z = round(gs$lower$bound, 3),
  Efficacy_p = round(2 * pnorm(-abs(gs$upper$bound)), 4)
) |>
  kable(
    col.names = c("Analysis", "Information Fraction", 
                  "N per Arm", "Total N",
                  "Efficacy Boundary (Z)", 
                  "Futility Boundary (Z)",
                  "Efficacy Boundary (p)"),
    caption = "Group Sequential Design: Stopping Boundaries and Sample Sizes"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = TRUE
  )

```

### Comparison: Fixed vs Sequential Design

```{r}

#| label: fixed-vs-sequential

tibble::tibble(
  Design = c("Fixed Sample", "Group Sequential (O'Brien-Fleming)"),
  N_per_arm = c(1524, ceiling(max(gs$n.I) / 2)),
  N_total = c(3048, ceiling(max(gs$n.I))),
  N_interim_looks = c(0, 3),
  Alpha_spent_interim = c("N/A", 
    paste0(percent(sum(gs$upper$spend[1:3]), accuracy = 0.01))),
  Expected_N_under_H1 = c(3048, 
    ceiling(gs$en[2]))
) |>
  mutate(
    N_per_arm = comma(N_per_arm),
    N_total = comma(N_total),
    Expected_N_under_H1 = comma(Expected_N_under_H1)
  ) |>
  kable(
    col.names = c("Design", "Max N per Arm", "Max Total N", 
                  "Interim Looks", "Alpha Spent at Interims",
                  "Expected N if H1 True"),
    caption = "Fixed Sample vs Group Sequential Design Comparison"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = TRUE
  )

```

### Interpretation

The group sequential design offers two key advantages over the fixed sample 
approach:

**Early stopping for efficacy:** If the intervention effect is substantially 
larger than the 3pp MDE, the trial can stop at Interim 1, 2, or 3, 
reducing expected enrollment duration. Under the alternative hypothesis, 
the expected sample size is smaller than the fixed sample maximum:
2,505 vs fixed sample 3,048 — the sequential design saves ~540 patients in expectation when the alternative is true.

**Controlled Type I error:** The O'Brien-Fleming spending function allocates 
very little alpha to early interim analyses — the first interim boundary 
is conservative, preserving most of the alpha budget for the final analysis. 
This means the trial will not stop early unless the evidence is overwhelming, 
reducing the risk of false positives from repeated testing.

**Practical implication for the deployment trial:** The group sequential 
design is strongly preferred over a fixed sample design given the 4-year 
enrollment horizon. Even modest early stopping at Interim 2 or 3 would 
meaningfully reduce operational burden and time to clinical impact.


## Section 4: Mock Pre-Registration

The following documents a structured pre-registration for the proposed trial, 
following the AsPredicted.org and OSF pre-registration frameworks adapted for 
a clinical RCT context. Pre-registration constrains researcher degrees of 
freedom and distinguishes confirmatory from exploratory analyses.

---

### Trial Title

**PREDICT-DM: A Randomized Controlled Trial of Machine Learning-Based 
Readmission Risk Scoring to Reduce 30-Day Hospital Readmission in Diabetic 
Inpatients**

---

### Primary Research Question

Among diabetic inpatients flagged as high-risk by a validated machine 
learning readmission risk model (predicted probability ≥ 10% (tau)), 
does exposure to an evidence-based intervention bundle triggered by the risk 
score reduce 30-day hospital readmission rates compared to standard care?

---

### Design Overview

```{r}

#| label: prereg-design

tibble::tibble(
  Parameter = c(
    "Trial design",
    "Randomization unit",
    "Treatment arms",
    "Enrollment criteria",
    "Primary outcome",
    "Secondary outcomes",
    "Follow-up period",
    "Blinding"
  ),
  Specification = c(
    "Parallel-group RCT with group sequential monitoring",
    "Individual patient encounter",
    "Intervention bundle vs. standard care (1:1 allocation)",
    "Adult diabetic inpatients with predicted readmission probability ≥ 10% at discharge planning",
    "30-day all-cause hospital readmission",
    "7-day readmission rate; emergency department visits within 30 days; length of stay",
    "30 days post-discharge",
    "Single-blind: patients and care teams unblinded; outcome assessors blinded"
  )
) |>
  kable(caption = "Trial Design Parameters") |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = TRUE
  )

```

---

### Primary Estimand

The primary estimand is the **Average Treatment Effect among flagged patients (ATT)** — 
the absolute risk reduction in 30-day readmission among patients 
with predicted readmission probability ≥ τ = 0.1 who receive the 
intervention bundle versus standard care:

$$ATT = E[Y(1) - Y(0) \mid \hat{p} \geq 0.1]$$

The estimand is defined on the **intention-to-treat (ITT)** population — 
all randomized patients regardless of intervention compliance.

---

### Sample Size Justification

```{r}

#| label: prereg-sample-size

tibble::tibble(
  Parameter = c(
    "Baseline readmission rate (control arm)",
    "Minimum detectable effect (ARR)",
    "Relative risk reduction",
    "Type I error rate (two-sided)",
    "Statistical power",
    "Required N per arm",
    "Required total N",
    "Sequential design",
    "Number of analyses",
    "Alpha spending function",
    "Expected N under H1",
    "Enrollment duration (single site)"
  ),
  Value = c(
    "11.2%",
    "3 perc pts",
    "26.8%",
    "0.05",
    "80%",
    "1,524",
    "3,048",
    "O'Brien-Fleming group sequential",
    "4 (3 interim + 1 final, annually)",
    "Lan-DeMets O'Brien-Fleming",
    "~2,505 (savings of ~543 vs fixed)",
    "~4 years (single site); ~12-16 months (multi-site, 3-4 hospitals)"
  )
) |>
  kable(caption = "Sample Size Justification") |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )

```

---

### Randomization Procedure

Patients flagged by the risk model at discharge planning will be randomized 
1:1 to intervention or control using **stratified block randomization** with 
stratification by:

- Primary diagnosis group (Diabetes vs. Circulatory vs. Other)
- Admission type (Emergency vs. Elective)
- Site (for multi-site implementation)

Block sizes will be randomly varied (4 and 6) and allocation concealed via 
a centralized web-based randomization system to prevent selection bias.

---

### Intervention Bundle Specification

```{r}

#| label: prereg-intervention

tibble::tibble(
  Component = c(
    "Enhanced discharge planning",
    "Expedited follow-up",
    "Transitional care enrollment",
    "Pharmacist reconciliation",
    "Targeted care gap closure"
  ),
  Timing = c(
    "24 hours before discharge",
    "Scheduled before discharge order signed",
    "Assigned within 2 hours of randomization",
    "Completed before discharge",
    "Completed before discharge"
  ),
  Responsible_Party = c(
    "Social work + care coordinator",
    "Scheduling system + attending physician",
    "Nurse navigator",
    "Clinical pharmacist",
    "Attending physician + care coordinator"
  )
) |>
  kable(
    col.names = c("Component", "Timing", "Responsible Party"),
    caption = "Intervention Bundle Specification"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = TRUE
  )

```

---

### Statistical Analysis Plan

**Primary analysis:**
Logistic regression of 30-day readmission on treatment arm indicator, 
adjusted for stratification variables (primary diagnosis group, admission 
type, site). Reported as odds ratio with 95% confidence interval and 
absolute risk difference with 95% CI.

**Sequential monitoring:**
Interim analyses conducted at approximately equal information fractions 
(25%, 50%, 75%, 100%) using O'Brien-Fleming spending function. The 
Data Safety Monitoring Board (DSMB) will review unblinded results at 
each interim. Stopping for efficacy requires crossing the pre-specified 
Z boundary. Stopping for futility is advisory, not binding.

**Secondary analyses:**

- Subgroup analyses by primary diagnosis group, age group, and HbA1c 
  testing status — all pre-specified, all reported with interaction 
  p-values, none powered for subgroup effects

- Sensitivity analysis excluding patients with protocol deviations

- Per-protocol analysis as sensitivity to ITT

**Handling of missing outcomes:**
Multiple imputation under missing-at-random (MAR) assumption for patients 
lost to follow-up. Sensitivity analysis under missing-not-at-random (MNAR) 
assumption using tipping point analysis.

---

### Stopping Rules

```{r}

#| label: prereg-stopping

tibble::tibble(
  Analysis = c("Interim 1 (Year 1)", "Interim 2 (Year 2)", 
               "Interim 3 (Year 3)", "Final (Year 4)"),
  Information_Fraction = round(gs$timing, 3),
  Efficacy_Z = round(gs$upper$bound, 3),
  Efficacy_p = round(2 * pnorm(-abs(gs$upper$bound)), 4),
  Action = c(
    "Stop for overwhelming efficacy only",
    "Stop for efficacy or advisory futility review",
    "Stop for efficacy or advisory futility review",
    "Final decision — stop regardless"
  )
) |>
  kable(
    col.names = c("Analysis", "Information Fraction",
                  "Efficacy Boundary (Z)", "Efficacy Boundary (p)",
                  "DSMB Action"),
    caption = "Pre-Specified Stopping Rules"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = TRUE
  )

```

---

### Deviations from Pre-Registration

Any analyses not specified above would be clearly labeled as **exploratory** 
in a final manuscript. The threshold τ = 0.1 and the MDE of 3 percentage 
points are fixed and may not be revised after enrollment begins without DSMB 
approval and protocol amendment. Model updating (retraining the risk model 
on new data) would not permitted during the trial period.

---

### Limitations and Threats to Validity

```{r}

#| label: prereg-limitations

tibble::tibble(
  Threat = c(
    "Unmeasured confounding",
    "Contamination",
    "Temporal outcome drift",
    "Model degradation",
    "Enrollment feasibility"
  ),
  Description = c(
    "Residual severity confounding not captured by observed covariates",
    "Control arm patients may receive elements of intervention bundle informally",
    "Readmission rates observed to decline over 1999-2008 study period; similar drift may occur during trial",
    "Risk model performance may degrade over time as patient population evolves",
    "4-year single-site enrollment horizon is at upper bound of operational feasibility"
  ),
  Mitigation = c(
    "Randomization eliminates confounding within flagged population; E-value sensitivity analysis planned",
    "Cluster randomization by ward or unit as alternative design if contamination detected",
    "Stratified randomization by enrollment period; temporal trend included as covariate in primary model",
    "Scheduled model performance review at each interim analysis; retraining protocol pre-specified",
    "Multi-site expansion pre-planned; adaptive sample size re-estimation at Interim 2 if enrollment lags"
  )
) |>
  kable(
    col.names = c("Threat", "Description", "Mitigation"),
    caption = "Threats to Validity and Mitigations"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = TRUE
  )

```